Safe Transferable Regions

### 2018-06-11 paper/broom.tex
Camera-Ready first cut

\input{macros}

%%%%%%%%%%%%%%%% OUR PRELIM STUFF ENDS %%%%%%%%%%%%%%%%%%%%

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Safe Transferable Regions}

@@

\acknowledgements{We are grateful to Kapil Vaswani, Dimitrios
Vytiniotis, Michael Isard, and Steven Hand for their valuable inputs
during the initial stages of this project. We would also like to thank
the anonymous reviewers for their careful scrutiny and insightful
comments that helped improve the quality of this paper.}%optional

@@

\begin{abstract}
\input{abstract}
\end{abstract}

\input{introduction}

\input{overview2}

\input{type-system}

\renewcommand{\rgn}{\pi}
\input{type-inference}
\input{misc}
\renewcommand{\rgn}{r}

\input{implementation}

\input{related}

\bibliography{broom}

>>>>

%--------------------------------------------------------------------
%Formatting Commands
%--------------------------------------------------------------------
\newcommand{\C}[1]{\code{#1}}
\renewcommand{\k}[1]{\mathtt{#1}} % Code font in math mode.
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\conj}{\wedge}
\newcommand{\disj}{\vee}
\newcommand{\tuplee}[1]{\langle #1 \rangle}
\newcommand{\ALT}{~\mid~}
\newcommand{\lamb}[2]{\lambda #1.\,#2}
\newcommand{\spc}[0]{\quad}
\newcommand{\redsto}[3]{#1 \vdash #2 \longrightarrow #3}
\newcommand{\rredsto}[0]{\rightsquigarrow_r }
\newcommand{\evalsto}[0]{\longrightarrow }
\newcommand{\env}[0]{\Gamma}
\newcommand{\tywf}[2]{#1\,\vdash\,#2 \; \texttt{ok}}
\newcommand{\fgjtywf}[2]{#1\,\Vdash\,#2 \; \texttt{ok}}
\newcommand{\isvalid}[2]{#1\,\vdash\,#2}
\newcommand{\hastyp}[3]{#1\,\vdash\,#2:#3}
\newcommand{\hasfgjtyp}[3]{#1\,\Vdash^{}\,#2\,:\,#3}
\newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
\newcommand{\fgjsubtyp}[3]{#1\,\Vdash\,#2 <: #3}
\newcommand{\msentails}[2]{#1\,\models\,#2}
\newcommand{\thesemof}[1]{ \llbracket #1 \rrbracket}
\newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
\newcommand{\absof}[1]{\thesemof{#1}}
\newcommand{\broom}{{\sc Broom}\xspace}
\newcommand{\name}{{\sc Broom}\xspace}
\newcommand{\namec}{{\sc Broomc}\xspace}
\newcommand{\fbname}{{\sc Featherweight Broom}\xspace}
\newcommand{\FB}{{\sc FB}\xspace}
\newcommand{\csharp}{C\#~}
\newcommand{\outlives}{\succeq}
\newcommand{\underlives}{\preceq}
\newcommand{\coloneqq}{::=}
\newcommand{\M}[1]{\mathtt{#1}}
\newcommand{\ObjZ}{\C{Object}}
\newcommand{\RgnZ}{\C{Region}}
\newcommand{\thisZ}{\C{this}}
\newcommand{\superZ}{\C{super}}
\newcommand{\unitZ}{\C{unit}}
\newcommand{\FuncZ}{\C{Func}}
\newcommand{\unitval}{\C{()}}
\newcommand{\inang}[1]{\langle #1 \rangle}
\newcommand{\rhoalloc}{\rho^a}
\newcommand{\rhobar}{\bar{\rho}}
\newcommand{\rhoallocm}{\rho^a_m}
\newcommand{\rhobarm}{\bar{\rho_m}}
\newcommand{\ralloc}{\pi^a}
\newcommand{\rgn}{r}
\newcommand{\rbar}{\bar{\rgn}}
\newcommand{\tbar}{\bar{T}}
\newcommand{\taubar}{\bar{\tau}}
\newcommand{\xbar}{\bar{x}}
\newcommand{\vbar}{\bar{v}}
\newcommand{\ebar}{\bar{e}}
\newcommand{\extends}{\triangleleft}
\newcommand{\rextends}{\blacktriangleleft}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\angR}{\inang{\rhobar \,|\, \phi}}
\newcommand{\angT}{\inang{\tbar}}
\newcommand{\tyvar}{a}
\newcommand{\tyvarb}{b}
\newcommand{\angAlpha}{\inang{\bar{\tyvar} \extends \bar{\fgjN}}}
\newcommand{\fgjN}{K}
\newcommand{\fbN}{N}
\newcommand{\headerOf}[1]{\C{class}\; #1\angAlpha\angR \extends \fbN}
\newcommand{\substFn}{\mathcal{S}}
% \newcommand{\regionSubstFn}{\mathcal{R}}
% \newcommand{\predSubstFn}{\mathcal{P}}
% \newcommand{\regionSubstFn}{\soln_V}
% \newcommand{\predSubstFn}{\soln_P}
\newcommand{\regionConstants}{R}
\newcommand{\regionVars}{V}
\newcommand{\predVars}{P}
\newcommand{\constraintSet}{C}
\newcommand{\regionDeltaMap}{{\Delta_R}}
\newcommand{\predDeltaMap}{{\Delta_P}}
\newcommand{\mang}{\inang{\bar{\rho_m} \,|\, \phi_m}}
\newcommand{\rhoset}{\Sigma}
\newcommand{\rhomap}{\Sigma}
\newcommand{\rhoenv}{\Delta}
\newcommand{\aenv}{\Theta}
\newcommand{\phicx}{\Phi}
\newcommand{\phictxt}{\phi_{cx}}
\newcommand{\phicstr}{\phi_{cs}}
\newcommand{\phisol}{\phi_{sol}}
\newcommand{\subtypcx}{\rhoenv,\aenv,\phicx}
\newcommand{\A}{{\mathcal{A}}}
\newcommand{\exptycx}[2]{\A,#1,#2}
\newcommand{\void}{\C{void}}
\newcommand{\functy}[3]{\FuncZ\inang{#1}\inang{#2,#3}}
\newcommand{\stmtsemcx}{\A,\env}
\newcommand{\stmtsemcxp}{\A',\env'}
\newcommand{\stmtsemcxpp}{\A'',\env''}
\newcommand{\sredsto}{\Rightarrow}
\newcommand{\stmtsem}[4]{#1 \, \vdash \, #2 \,/\, #3 \sredsto #4}
\newcommand{\okin}[2]{#1 \; \texttt{ok} \; \texttt{in} \; #2}
\newcommand{\lambdaexp}[4]{\lambda@#1\inang{#2}(#3). #4}
\newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
\newcommand{\letexp}[3]{\C{let}\;#1\,=\,#2\;\C{in}\;#3}
\newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
\newcommand{\open}[4]{\C{open}\;#1\;\C{as}\; #3@#2\;\C{in}\;#4}
\newcommand{\opened}[3]{\C{opened}\;#1(#2)\;\C{in}\;#3}
\newcommand{\letd}[2]{\C{letd}\;#1\;\C{in}\;#2}
\newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
\newcommand{\invalidexn}{\bot}
\newcommand{\toprgn}{\pi_{\top}}
\newcommand{\csolvestar}{{\sc CSolve*}\xspace}
\newcommand{\elabExpr}{{\sf elabExpr}}
\newcommand{\fields}{{\sf fields}}
\newcommand{\mtype}{{\sf mtype}}
\newcommand{\mbody}{{\sf mbody}}
\newcommand{\bound}{{\sf bound}}
\newcommand{\fresh}{{\sf fresh}}
\newcommand{\typeOk}{{\sf typeOk}}
\newcommand{\subtypeOk}{{\sf subtypeOk}}
\newcommand{\templateTy}{{\sf templateTy}}
\newcommand{\allocRgn}{{\sf allocRgn}}
\newcommand{\shape}{{\sf shape}}
\newcommand{\ctype}{{\sf ctype}}
\newcommand{\override}{{\sf override}}
%\newcommand{\smallskip}{\vspace*{0.1in}}
\newcommand{\elabMeth}{{\sf elabMeth}}
\newcommand{\frv}{{\sf frv}}
\newcommand{\rv}{{\sf RVars}}
\newcommand{\solve}{{\sf solve}}
\newcommand{\valuee}{{\sf value}}
\newcommand{\elabClass}{{\sf elabClass}}
\newcommand{\elabCons}{{\sf elabCons}}
\newcommand{\elabClassTable}{{\sf elabClassTable}}
\newcommand{\CLOSED}{\mathtt{C}}
\newcommand{\OPEN}{\mathtt{O}}
\newcommand{\XFERRED}{{\color{blue}\times}}
\newcommand{\LIVE}{{\color{ForestGreen}\blacksquare}}
\newcommand{\SLIVE}{\LIVE}
\newcommand{\TLIVE}{\LIVE}
% \newcommand{\SLIVE}{{\color{ForestGreen}\widehat{\blacksquare}}}
% \newcommand{\TLIVE}{{\color{ForestGreen}\overrightarrow{\blacksquare}}}
%\newcommand{\FREE}{{\color{blue}\square}}
% \newcommand{\USED}{{\color{blue}\overrightarrow{\blacksquare}}}
\newcommand{\USED}{{\color{blue} \square}}
\newcommand{\transfer}{{\sf transfer}}
\newcommand{\CSP}{(C\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi}, \bar{\rhoenv})}
\newcommand{\csp}{(c\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi},\bar{\rhoenv})}
\newcommand{\Phicx}{\Phi_{cx}}
\newcommand{\Phics}{\Phi_{cs}}
\newcommand{\tynwf}[2]{#1\,\nvdash\,#2 \; \texttt{ok}}
\newcommand{\csid}[1]{\lbrack \mathbf{c_{#1}} \rbrack:\;}
\newcommand{\qqquad}{\quad\quad\quad}
\newcommand{\emptyA}[0]{(\Delta,\cdot,\phicx)}
\newcommand{\emptyASigp}[0]{(\Delta,\cdot,\phicx)}
\newcommand{\RgnZT}[1]{\RgnZ\inang{T}\inang{#1}}
\newcommand{\BZ}[0]{B\inang{\tbar}}
\newcommand{\BZT}[1]{B\inang{\tbar}\inang{#1}}
\newcommand{\redstoo}[3]{#2 \longrightarrow #3} %\Delta ignored
\newcommand{\redstocup}[3]{\redstoo{\rhoenv \cup \{#1\}}{#2}{#3}}
\newcommand{\loc}{\mathtt{l}}
\newcommand{\locbar}{\overline{\loc}}
\newcommand{\isLive}{{\sf isLive}}
\newcommand{\mem}{\Sigma}
\newcommand{\status}{s}
\newcommand{\toploc}{\loc_{\top}}
\newcommand{\consistent}{{\sf consistent}}

%Rules
%-----
\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
\newcommand{\rulee}[1]{\textrm{\sc {#1}}}
\newcommand{\ilrulelabel}[1]{{\sc #1}}
\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
                           {\begin{array}{c}#2\end{array}}}

% Math mode
%-----------
\newenvironment{nop}{}{}
\newenvironment{smathpar}{
\begin{nop}\small\begin{mathpar}}{
\end{mathpar}\end{nop}\ignorespacesafterend}

% Theorems Environment
%----------------------
% \newtheorem{theorem}{Theorem}[section]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{proposition}[theorem]{Proposition}
% \newtheorem{corollary}[theorem]{Corollary}

% \newenvironment{proof}[1][Proof]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
% \newenvironment{definition}[1][Definition]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
% \newenvironment{example}[1][Example]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
% \newenvironment{remark}[1][Remark]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

% \newcommand{\qed}{\nobreak \ifvmode \relax \else
      % \ifdim\lastskip<1.5em \hskip-\lastskip
      % \hskip1.5em plus0em minus0.5em \fi \nobreak
      % \vrule height0.75em width0.5em depth0.25em\fi}

\usepackage{listings}
\usepackage{subcaption}
\usepackage{stmaryrd}
\usepackage[skip=5pt]{caption}

\definecolor{Burgundy}{rgb}{0.5, 0.0, 0.13}
\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
\definecolor{ForestGreen}{rgb}{0.13, 0.55, 0.13}
\definecolor{Bubbles}{rgb}{0.91, 1.0, 1.0}
\definecolor{LighterGray}{rgb}{0.93, 0.93, 0.93}

\newcommand{\lstjava}{\lstset{ %
language=Java, % choose the language of the code
backgroundcolor=\color{LighterGray},   
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Burgundy}\bf,
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={var, foreach, unit, letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
}}
\lstnewenvironment{codejava}
    { % \centering
			\lstjava
      \lstset{mathescape=true}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

\lstnewenvironment{numcodejava}
    { % \centering
			\lstjava
      \lstset{numbers=left,firstnumber=1}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

%--------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%% OUR PRELIM STUFF ENDS %%%%%%%%%%%%%%%%%%%%%%
%--------------------------------------------------------------------

\title{Safe Transferable Regions}

@@

\acknowledgements{We are grateful to Kapil Vaswani, Dimitrios
Vytiniotis, Michael Isard, and Steven Hand for their valuable inputs
during the initial stages of this project. We would also like to thank
multiple anonymous reviewers for their careful scrutiny and feedback at
various stages of this work.}%optional

@@

%--------------------------------------------------------------------
%                           Abstract
%--------------------------------------------------------------------
\begin{abstract}
There is an increasing interest in alternative memory management
schemes that seek to combine the convenience of garbage collection and
the performance of manual memory management in a single language
framework.  Unfortunately, ensuring safety in presence of manual
memory management remains as great a challenge as ever. In this paper,
we present a C\#-like object-oriented language called {\sc
Broom}\xspace that uses a combination of region type system and
lightweight runtime checks to enforce safety in presence of
user-managed memory regions called \emph{transferable regions}. Unsafe
transferable regions have been previously used to contain the latency
due to unbounded GC pauses. Our approach shows that it is possible to
restore safety without compromising on the benefits of transferable
regions.  We prove the type safety of {\sc Broom}\xspace in a formal
framework that includes its C\#-inspired features, such as
higher-order functions and generics. We complement our type system
with a type inference algorithm, which eliminates the need for
programmers to write region annotations on types. The inference
algorithm has been proven sound and relatively complete. We describe a
prototype implementation of the inference algorithm, and our
experience of using it to enforce memory safety in dataflow programs.
\end{abstract}

%--------------------------------------------------------------------
%                           Introduction
%--------------------------------------------------------------------
\newcommand{\TODO}[1]{\textbf{TODO: #1}} \newcommand{\eg}{\emph{e.g.}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\mypara}[1]{\textbf{#1}}

\section{Introduction} \label{sec:introduction}

% This paper presents a \emph{safe} and \emph{efficient} region-based approach
% to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.

Computations performed by a network of concurrent communicating actors
often involve data transfer between producers and consumers. Consider
for example the \C{SELECT} query operator shown in
Fig.~\ref{fig:motivating-eg}, which functions as an actor in a
dataflow computation involving a network of other such query
operators. \C{SELECT} receives a stream of input messages, each
associated with a time window $t$, processed by method
\texttt{onReceive}.  Each input message contains a list of inputs,
each processed by applying a user-defined function to create a
corresponding output.  Multiple messages with the same timestamp are
permitted and messages with different timestamps may arrive out of
order.  An invocation of method \texttt{OnNotify} indicates that no
more input messages with a timestamp $t$ will be subsequently
delivered.  At this point, the operator completes the processing for
time window $t$ and sends a corresponding output message to its
successor.

\mypara{Performance.}
When the transferred data structures are large, as is the case with many
streaming big-data analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.  Further, in a distributed
dataflow system, the GC pause at one node can have a cascading adverse
effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
a GC pause at an upstream actor can block downstream actors that are waiting for
messages.  However, much of the GC overhead results from the
collector performing avoidable or unproductive work.  For the example
in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
\C{map}, although its objects cannot be collected until a
suitable timing message arrives.
There has been increasing interest in off-heap memory management for better performance
in the context of several  systems and languages (e.g.,
Spark~\cite{SPARK:2015}, Scala~\cite{ScalaOffHeap}, Rust).
Several systems (e.g.,~\cite{TRILL:2014}) resort to using ``free object pools'' to partially alleviate this problem, in
the absence of language support.

\begin{figure}[t!]
\begin{numcodejava}
class SelectVertex<TIn, TOut> {
  Func<TIn, TOut> selector;
  Dictionary<Time, List<TOut>> map;
  ...
  void onReceive(Time t, List<TIn> inList) {
    if (!map.ContainsKey(t)) map[t] = new List<TOut>();
    foreach (TIn input in inList) {
      TOut output = selector(input);
      map[t].add(output); } }
  void onNotify(Time t) {
     List<TOut> outList = map[t];
     map.Remove(t);
     transfer(successorId, t, outList); }
}
\end{numcodejava}
\caption{\C{SELECT} dataflow operator}
\label{fig:motivating-eg}
\vspace*{-0.2in}
\end{figure}

\begin{figure}
\centering
\subcaptionbox {
  Before transfer.
  \label{fig:after-transfer}
} [
  0.35\columnwidth
] {
  \includegraphics[scale=0.36]{producer-heap}
}
\subcaptionbox {
  After transfer.
  \label{fig:before-transfer}
} {
  \includegraphics[scale=0.36]{producer-consumer-heap}
}

\caption{References in and out of a transferable region ($R$) become
invalid after transfer.}
\label{fig:transfer-violations}
\end{figure}

\mypara{Safety.}
Several of these systems are designed to allow the producer and
consumer to execute in the same address space or in different address
spaces (as determined by the compiler and runtime).  We wish to ensure
memory safety in the presence of such transferred data.  A transfer
operation, with no additional checks, may cause memory safety
violations, both at the producer of the data structure, and its
(possibly remote) consumer. At the producer, any existing references
into the transferred data structure become invalid post transfer. If
the data structure contains references to objects outside (the
transferred data), then such references become invalid in the context
of the consumer. Both scenarios are depicted in
Fig.~\ref{fig:transfer-violations}.  While references from within the
transferable data to outside can be disallowed in the interest of
safety, references into transferable data needs to be allowed before
the data is transferred. Allowing such references is crucial, as any
non-trivial program creates temporary references to the internal
objects of a data-structure.

\mypara{Safe Transferable Regions.}
In this paper, we present a language-based approach to cleanly
encapsulate transferable data and a safe and efficient implementation
of such data based on regions. 
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

The use of regions alleviates the performance concern related to
garbage collection described earlier. (It also enables more efficient
serialization/deserialization of data-structures across address spaces,
which is also a significant performance bottleneck in such systems.)
However, manual region memory management introduces memory safety
issues such as dangling pointers.
We present a single type system that simultaneously addresses
this safety concern, as well as those described above related to
the use of \emph{transferable data}.

Safe region-based memory management using types was
pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
use only lexically scoped regions.  At runtime, the set of all regions
(existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two
regions whose lifetimes overlap, with neither one's lifetime contained
within the other.  Unfortunately, the data structures in the above
example do not satisfy this restriction (as the output messages for
multiple time windows may be simultaneously live, without any
containment relation between their lifetimes).  We refer to regions
with lexically scoped lifetimes as \emph{stack regions} and to regions
that do not have such a lexically scoped region as \emph{dynamic}
regions.

Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
that can be safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions.
%
Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
that combine linear types and regions to support dynamic regions.
Unlike the prior work, we ensure memory safety in the presence of
transferable regions through a \emph{combination of a static
typing discipline and lightweight runtime checks in place of linear types}.

\mypara{Language.}
The cornerstone of this approach is an \C{open} lexical block for
transferable regions, that ``opens'' a transferable region and
guarantees that the region won't be transferred/freed while it is
open. By nesting a Tofte-Talpin style \C{letregion} lexical block,
that delimits the lifetime of a stack region, inside an \C{open}
lexical block for a transferable region, we can guarantee that the
transferable region will remain live as long as the stack region is
live. We say that the former \emph{outlives} the latter, and any
references from the stack region to the transferable region are
therefore safe. This is particularly useful, since such stack regions
serve as temporary working storage while working with the (opened)
transferable regions.

By controlling the outlives relationships between various
regions, we only allow safe cross-region references, while prohibiting
unsafe ones. In the above example, an outlives relationship
\emph{from} the stack region \emph{to} the transferable region means
that the references in that direction are allowed, but not the
references in the opposite direction. In contrast, if an \C{open}
block of a transferable region $R_0$ is nested inside an \C{open}
block of another transferable region $R_1$, we do not establish any
outlives relationships, thus declaring our intention to not allow any
cross-region references between $R_0$ and $R_1$.  Finally, we observe
that outlives relationships are established based on the lexical
structure of the program, hence a static type system can enforce them
effectively. By assigning region types to objects, which capture the
regions such objects are allocated in, and by maintaining outlives
relationships between various regions, we can statically decide the
safety of all references in the program.

\mypara{Type System.}
Formally, our type system introduces \emph{region parameters} (for both
classes and methods) and uses constrained parametric polymorphism
over these parameters, where the constraints capture outlives constraints
between the region parameters.
The type system may be seen as a form of ownership type system~\cite{OwnershipSurvey},
with a region being the owner of all objects allocated in that region.

\mypara{Lightweight Runtime Checks.}
Ensuring memory safety using this approach requires ensuring
that the use of transferable regions satisfies certain temporal properties.
Firstly, a transferable region should not be transferred/freed
inside an open block of that region (i.e, while it is still open).
Secondly, a transferred/freed region should not be opened. These are
typestate invariants on the transferable region objects, which are
hard to enforce statically in the presence of unrestricted
aliasing. Techniques like linear types and unique pointers can be used
to restrict aliasing, but the constraints they impose are often hard
to program around. We therefore enforce typestate invariants at
runtime via lightweight checks. In particular, we define an acceptable
state transition discipline for transferable regions
(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
transition of a transferable region (\eg, from \emph{open} state to
\emph{freed} state) is valid or not. The check is lightweight since it
only involves checking a single tag that captures the current state.
We believe that this is a reasonable choice since regions are
coarse-grained objects manipulated infrequently, when compared to the
fine-grained objects that are present inside these regions, for which
safety is enforced statically.

\mypara{Type Inference.}
One of the key contributions of this paper is a type inference
algorithm that eliminates the need for users to write region type
annotations.  The users write programs in the underlying language that
provides primitives for the users to manipulate regions (create, open,
free, transfer) and to allocate objects in regions, but has no region
type annotations.

Our inference algorithm proceeds in three stages: in the first stage,
we elaborate the program by introducing region parameters (for classes
and methods); in the second stage we generate a set of constraints
that must be satisfied for the program to type check; in the third
stage, we solve the set of constraints, inferring the preconditions of
all classes and methods (in terms of the expected outlives-constraints
between the region parameters).  We show that the algorithm is sound,
and the stages of constraint generation and solving are complete
(i.e., the algorithm is complete relative to the first stage of
elaboration).
% (using an interprocedural fixed point computation).

\mypara{Evaluation.}
Our work was inspired by~\cite{Broom:HotOS}, which presents evidence
that realistic programs can be written using transferable regions and
that this can yield significant performance improvements.
While~\cite{Broom:HotOS} addresses the engineering challenges in
extending a managed runtime with transferable regions, it adopts an ad
hoc approach insofar as language design is concerned, exposing the
region functionality through an unsafe API, and in the process losing
the safety guarantees. Our contribution is an alternative approach
that is grounded in sound theory and restores the language safety
guarantees. The utility of our approach is demonstrated by a prototype
implementation of our type inference algorithm that was able to
identify unsafe memory accesses among the benchmarks extracted
from~\cite{Broom:HotOS}.

\mypara{Contributions.}
The paper makes the following contributions:

\begin{itemize} 
  \item We present \broom, a \csharp-like typed object-oriented
  language that supports programmer-managed memory regions. \broom{}
  extends its core language, which includes \emph{lambdas}
  (higher-order functions) and \emph{generics} (parametric
  polymorphism), with constructs to create, manage and free static and
  transferable memory regions. Transferable regions are first-class
  values in \broom.

\item \broom{} is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  provided that certain typestate invariants on regions hold.  The
  latter invariants are enforced via simple runtime checks.

  \item We define an operational semantics for \broom, and a type
  safety result that clearly defines and proves safety guarantees
  described above.

  \item We describe a region type inference algorithm for \broom that
  (a). completely eliminates the need to annotate \broom programs with
  region types, and (b). enables seamless interoperability between
  region-aware \broom programs and legacy standard library code that
  is region-oblivious. Our inference algorithm is based on a novel
  constraint solver that performs abduction in a partial-order
  constraint domain to infer weakest solutions to recursive
  constraints.

\item We establish the soundness and relative completeness of the type
inference algorithm.

  \item We describe an implementation of \broom frontend in OCaml,
  along with case studies where the region type system was able to
  identify unsafe memory accesses statically.

\end{itemize}

%--------------------------------------------------------------------
%                             Overview
%--------------------------------------------------------------------
\newcommand{\COMMENT}[1]{}

\section{An Informal Overview of \name} \label{sec:overview}

\name enriches a simple object-oriented language (supporting
parametric polymorphism and lambdas) with a set of region-specific
constructs.  In this section, we present an informal overview of these
region-specific constructs.

\subsection{Using Regions in \name}
\label{sec:alloc-ctxt}

\mypara{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
creates a new stack region, with a static identifier \C{R}, whose
scope is restricted to the statement \C{S}. The semantics of
\C{letregion} is similar to Tofte and Talpin~\cite{tofte94}'s
\C{letregion} expression: objects can be allocated by \C{S} in the
newly created region while \C{R} is in scope, but the region and all
objects allocated within it are freed at the end of \C{S}.

\mypara{Object Allocation} The ``\C{new@R T()}'' construct creates
a new object of type \C{T} in the region \C{R}. The specification of
the allocation region \C{R} in this construct is optional.  At
runtime, \name maintains a stack of \emph{active} regions, and we
refer to the region at the top of the stack as the \emph{allocation
context}. The statement \C{new T()} allocates the newly created object
in the current allocation context.
%
This is important as it enables \name applications to use existing
region-oblivious C\# libraries, as explained soon.

\mypara{Transferable Regions} 
Transferable regions are first class values of \name: they are objects
of the class \C{Region}, they are created using the \C{new} keyword,
and can be passed as arguments, stored in data structures, and
returned from methods.  A transferable region is intended to
encapsulate a single data-structure, consisting of a collection of
objects with a distinguished root object of some type \C{T}, which we
refer to as the region's \emph{root} object.  The class \C{Region} is
parametric over the type \C{T} of this root object.

The \C{Region} constructor takes as a parameter a function that
constructs the root object: it creates a new region and invokes this
function, with the new region as the allocation context, to create the
root object of the region. The following code illustrates the
creation of a transferable region, whose root is an object of type
\C{A}.
% (\C{r}) of type \C{T}:
\begin{codejava} 
  Region<A> rgn = new Region<A>(() => new A())
\end{codejava} 
In the above code, \C{rgn} is called the \emph{handler} to the newly
created region, and is required to read the contents of the region, or
change its state. The class \C{Region} offers two methods: a \C{free}
method that deallocates the region (and all the objects allocated
within it), and \C{transfer} method that transfers the region to a
(possibly remote) consumer process.

\mypara{Open and Closed Regions} A transferable region must be
explicitly \emph{opened} using \name's \C{open} construct in order to
either read or update or allocate objects in the region.
Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
following: (a). It opens the transferable region handled by \C{rgn}
for allocation (i.e., makes it the current allocation context), (b).
binds the identifier \C{R} to this open region, and (c). initializes
the newly introduced local variable \C{v} to refer to the root object
of the region.  The \C{@R} part of the statement is optional and may
be omitted.  The \C{open} construct is intended to simplify the
problem of ensuring memory safety, as will be explained soon.  We
refer to a transferable region that has not been opened as a
\emph{closed} region. A transferable region can only be transferred or
freed when it is in closed state. The acceptable state transition
discipline over the lifetime of a transferable region is described
in Fig.~\ref{fig:region-fsm}. Enforcement of this discipline is done
at runtime.

Opening a region also makes it the current allocation context.
Thus, given a  C\# library function \C{f} (that makes no use of \name's region constructs),
opening a region \C{R} and invoking \C{f} has the effect that all objects created by this
invocation are allocated in the region \C{R}.

\mypara{Motivating Example}
Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name. The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient). On Line 7, we create a new region to
store the output for time \C{t}, initializing it to contain an empty
list. On Lines 8 and 9, we open the input region \C{inRgn} followed by
creating a new stack region \C{R0}. The temporary objects created by
the iteration on line 10, for example, will be allocated in the stack
region \C{R0} that lives just long enough.  We open the desired output
region on line 11, so that the new output objects created by the
invocation of \C{selector} on line 12 are allocated in the output
region.  Finally, the input region is freed on line 14. The output
region at \C{map[t]} stays as along as input messages with timestamp
\C{t} keep arriving. When the timing message for \C{t} arrives, the
\C{onNotify} method transfers the \C{outRgn} at \C{map[t]} to a
downstream actor.

\begin{figure}[t!]
\begin{numcodejava}
class SelectVertex<TIn, TOut> {
  Func<TIn, TOut> selector;
  Dictionary<Time, Region<List<TOut>>> map;
  ...
  void onReceive(Time t,Region<List<TIn>> inRgn){
    if (!map.ContainsKey(t))
       map[t] = new Region<List<TOut>> (() => new List<TOut>());
    open inRgn as inList {
      letregion R0 {
        foreach (TIn input in inList) {
          open map[t] as outList {
            TOut output = selector(input);
            outList.add(output); } } } }
    inRgn.free();
  }
  void onNotify(Time t) {
     Region<List<TOut>> outRgn = map[t];
     map.Remove(t);
     outRgn.transfer(successorId);
  }
}
\end{numcodejava}
\caption{\C{SELECT} dataflow operator in \name}
\label{fig:motivating-eg-in-broom}
\vspace*{-0.15in}
\end{figure}

\begin{figure} 
\includegraphics[scale=0.45]{region-fsm.png}
\caption{The lifetime of a dynamic (transferable) region in \name}
\label{fig:region-fsm} 
\vspace*{-0.1in} 
\end{figure}
